Databases are an effective and efficient method for storage and access of mass-spectrometry data

### 2025-06-10 manuscript/manuscript.Rmd
Numbering supplemental figures

```{r}
library(gt)
# data.frame(
#   filename=c("Smp_A", "Smp_A", "Smp_Z", "Smp_Z"),
#   scan_num=c(1,1,651501,651501),
#   rt=c(0.10, 0.10, 12.37, 12.37),
#   dt=c(),
#   mz=c(60.0452, 60.0532, 60.0456, 60.0531),
#   int=c(6618, 2657, 158084, 4673)
# ) %>%
#   gt() %>%
#   cols_label(filename=html("filename</br><small>Source file</small>"),
#              scan_num=html("scan_num</br><small>Scan number</small>"),
#              rt=html("rt</br><small>Retention time</small>"), 
#              mz=html("mz</br><small><emph>m/z</emph> ratio</small>"), 
#              int=html("int</br><small>Intensity</small>")) %>%
#   tab_row_group(label = "...millions of additional entries...", rows = rt>=mean(rt)) %>%
#   tab_row_group(label = "", rows = rt<=mean(rt)) %>%
#   tab_style(style = cell_text(align = "center"), locations = cells_row_groups()) %>%
#   opt_table_outline() %>%
#   tab_style(cell_borders(sides = c("left", "right")), locations = cells_body()) %>%
#   cols_align(align = "center") %>%
#   cols_move_to_start("filename") %>%
#   tab_options(row_group.padding = 3) %>%
#   tab_header(title = "Table: MS1") %>%
#   tab_style(style = cell_text(align = "left"), locations = cells_title()) %>%
#   gtsave(filename = "figures/ms1_table.png", vwidth=1000)
```

*Supplemental figure XX: [TBD] Additional possible extensions to the database schema for retention time correction, feature detection and correspondence, ion mobility, and imaging.*

@@

The simplest and most abundantly documented query was the extraction of a single spectrum. In many ways, this is the fundamental unit of mass spectrometry and thus many formats are highly optimized for its extraction into manipulatable data (Figure 2A and 2D). Here, we found that the mzML and mzMLb file types were consistently the slowest to parse and required multiple seconds, likely highlighting inefficiencies in the `pyteomics` package used to parse both file types. `pyopenms` also struggled to open and extract a specific scan, requiring several seconds due in large part to the expensive initiation function, after which requests were orders of magnitude faster (Supplemental Figure XX). It is also worth noting that while both of these packages provided rapid extraction of a *random* spectrum, a significant overhead was introduced by needing to scan through the file to find a *specific* spectrum by scan number. Scans are not always consecutive (e.g. during multiple MS experiments, polarity-switching runs, or when scans have been filtered out during conversion) and no metadata was obviously available that would have allowed using the index directly to a specific scan number.

@@

init_outside_gp / pyopenms_precomp_gp
ggsave("figures/supp_preload_fig.png", width = 6.5, height=5)
```

![](figures/supp_preload_fig.png)

*Supplemental figure XX: Timing information with and without pre-loading the object into Python prior to extraction. Panel A) shows the number of seconds required to extract a scan for the three mzML parsers plus the mzMLb, MZA, and mz5 file types. Panel B) shows the number of seconds required for pyopenms to perform each of the queries if the initial data load was removed from the timing count. Error bars show 3x replicate queries of 5x different targets.*

@@

Most methods were linear extrapolations of the single chromatogram numbers shown above as expected from a simple for loop, with notable exceptions for pyteomics (and thus the mzMLb format), the 2D peak method of pyopenms, and DuckDB (Figure 3). Pyteomics and pyopenms both had significant overhead upon initial load that resulted in faster query times afterward that performed better than a 1:1 extrapolation would expect, with pyopenms matching SQLite's speed after 10 chromatograms and Parquet's speed after 30. Interestingly, the methods that had a best-fit linear slope less than 1:1 also all had exponential fits, with performance at high chromatogram number actually worse than would be expected from a prediction fit to the timings for 1 and 3 chromatogram extractions (Supplemental figure XX).

```{r}
multichrom_times <- read_csv("data/multichrom_times.csv")
multichrom_baseline_df <- multichrom_times %>%
  mutate(method=str_replace(method, "pyopenms_2DPeak", "2DPeak")) %>%
  filter(n_chroms==1) %>%
  summarise(time=mean(time), .by=method) %>%
  mutate(method=factor(method, levels=method_names))
multichrom_times %>%
  mutate(method=str_replace(method, "pyopenms_2DPeak", "2DPeak")) %>%
  mutate(method=factor(method, levels=method_names)) %>%
  ggplot(aes(x=n_chroms, y=time, color=method, fill=method)) +
  geom_abline(aes(slope=1, intercept=log10(time)), data=multichrom_baseline_df) +
  geom_smooth(method="lm", formula="y~x", color="black") +
  geom_point(position = position_dodge(width=0.1), color="black", pch=21, size=3) +
  scale_y_log10(expand=expansion(0.1)) +
  scale_x_log10(breaks=unique(multichrom_times$n_chroms), expand=expansion(0.1)) +
  scale_fill_manual(breaks = method_names, values=method_colors, 
                    aesthetics = c("fill", "color")) +
  facet_wrap(~method, scales="free_y") +
  labs(x="Number of chromatograms extracted", y="Time required (seconds)") +
  theme_bw() +
  theme(legend.position = "none")
ggsave("figures/supp_multichrom_facets.png", width = 6.5, height=5)
```

![](figures/supp_multichrom_facets.png)

*Supplemental figure XX: Data from main text Figure 3 with independent y-axes showing scatter plots of the time required to extract multiple chromatograms using various methods on logarithmic axes. Best-fit linear models have been added for each method and are shown behind triplicate timing measurements, while the black line shows a 1:1 slope intercepting the first data point to show nonlinear behavior with additional chromatograms. Transparent intervals around each best-fit line show a single standard error of the mean.*

We also explored whether database queries could be improved via the use of either a unified query (single SQL statement with multiple OR clauses for each ion's *m/z* range) or a non-equi join between a peak table with *m/z* minimum and maximum columns (Supplemental Figure XX). SQLite and Parquet performed ~3-5 times faster with the unified query than with the loop method despite the necessity of and additional processing step for the looped query to correctly assign each data point to its original peak information. The opposite was true for DuckDB likely due to its optimized reader, with the unified query consistently outperformed by the non-equi join when 100 chromatograms were extracted.

```{r}
type_labels = c("<span style='color:#a41118'>For loop</span>", 
               "<span style='color:#f4bb23'>Unified query</span>",
               "<span style='color:#147C8F'>Non-equi join</span>")
read_csv("data/dbloopjoinuni_times.csv") %>%
  separate(method, into = c("method", "type")) %>%
  mutate(type=case_when(
    type=="join"~"Non-equi join",
    type=="loop"~"For loop",
    type=="unified"~"Unified query"
  )) %>%
  mutate(type=factor(type, levels=c("For loop", "Unified query", "Non-equi join"),
                     labels=type_labels)) %>%
  mutate(method=factor(method, levels=c("sqlite", "duckdb", "parquet"),
                       labels=c("SQLite", "DuckDB", "Parquet"))) %>%
  ggplot(aes(x=n_chroms, y=time, fill=type)) +
  geom_point(position=position_dodge(width = 0.1), color="black", size=3, pch=21) +
  facet_wrap(~method, nrow=1) +
  scale_y_log10() +
  scale_x_log10(breaks=c(1, 3, 10, 30, 100)) +
  scale_fill_manual(breaks=type_labels, values = c("#a41118", "#f4bb23", "#147C8F")) +
  labs(x="Number of chromatograms", y="Time required (seconds)", fill=NULL) +
  theme_bw() +
  theme(legend.position = "inside", legend.justification.inside = c(0.5, 0.95),
        legend.background = element_rect(color='black'),
        legend.text=element_markdown(),
        plot.title.position = "plot", 
        plot.caption.position =  "plot")
ggsave("figures/supp_dbloopjoinuni.png", width = 6.5, height=5)
```

![](figures/supp_dbloopjoinuni.png)

*Supplemental figure XX: Scatter showing time required to extract chromatograms from SQLite, DuckDB, and Parquet databases using different methods denoted by color. Ten replicates are shown for each data point.*

@@

Certain methods are clear winners for individual use cases but all of the existing formats failed to perform well at the full suite of exploratory data analysis tasks we attempted. Scan extraction is perhaps the most widely used query, especially for proteomics, and as a result has been extensively optimized. Here, MZA was blazingly fast thanks to the decision to index in the HDF5 file by scan number. The mzML files used here did have a precompiled index that should have made scan extraction highly efficient but this appeared to be mitigated for pyopenms and pyteomics due to their long initialization times, though pymzml performed very well at this task and the other two mzML methods were much faster after initialization (Supp Fig XX).

>>>>

```{r supp fig 1 gtables, eval=FALSE}
library(gt)
# data.frame(
#   filename=c("Smp_A", "Smp_A", "Smp_Z", "Smp_Z"),
#   scan_num=c(1,1,651501,651501),
#   rt=c(0.10, 0.10, 12.37, 12.37),
#   dt=c(),
#   mz=c(60.0452, 60.0532, 60.0456, 60.0531),
#   int=c(6618, 2657, 158084, 4673)
# ) %>%
#   gt() %>%
#   cols_label(filename=html("filename</br><small>Source file</small>"),
#              scan_num=html("scan_num</br><small>Scan number</small>"),
#              rt=html("rt</br><small>Retention time</small>"), 
#              mz=html("mz</br><small><emph>m/z</emph> ratio</small>"), 
#              int=html("int</br><small>Intensity</small>")) %>%
#   tab_row_group(label = "...millions of additional entries...", rows = rt>=mean(rt)) %>%
#   tab_row_group(label = "", rows = rt<=mean(rt)) %>%
#   tab_style(style = cell_text(align = "center"), locations = cells_row_groups()) %>%
#   opt_table_outline() %>%
#   tab_style(cell_borders(sides = c("left", "right")), locations = cells_body()) %>%
#   cols_align(align = "center") %>%
#   cols_move_to_start("filename") %>%
#   tab_options(row_group.padding = 3) %>%
#   tab_header(title = "Table: MS1") %>%
#   tab_style(style = cell_text(align = "left"), locations = cells_title()) %>%
#   gtsave(filename = "figures/ms1_table.png", vwidth=1000)
```

*Supplemental figure 1: [TBD] Additional possible extensions to the database schema for retention time correction, feature detection and correspondence, ion mobility, and imaging.*

@@

The simplest and most abundantly documented query was the extraction of a single spectrum. In many ways, this is the fundamental unit of mass spectrometry and thus many formats are highly optimized for its extraction into manipulatable data (Figure 2A and 2D). Here, we found that the mzML and mzMLb file types were consistently the slowest to parse and required multiple seconds, likely highlighting inefficiencies in the `pyteomics` package used to parse both file types. `pyopenms` also struggled to open and extract a specific scan, requiring several seconds due in large part to the expensive initiation function, after which requests were orders of magnitude faster (Supplemental Figure 2). It is also worth noting that while both of these packages provided rapid extraction of a *random* spectrum, a significant overhead was introduced by needing to scan through the file to find a *specific* spectrum by scan number. Scans are not always consecutive (e.g. during multiple MS experiments, polarity-switching runs, or when scans have been filtered out during conversion) and no metadata was obviously available that would have allowed using the index directly to a specific scan number.

@@

init_outside_gp / pyopenms_precomp_gp
ggsave("figures/supp_2_preload_fig.png", width = 6.5, height=5)
```

![](figures/supp_2_preload_fig.png)

*Supplemental figure 2: Timing information with and without pre-loading the object into Python prior to extraction. Panel A) shows the number of seconds required to extract a scan for the three mzML parsers plus the mzMLb, MZA, and mz5 file types. Panel B) shows the number of seconds required for pyopenms to perform each of the queries if the initial data load was removed from the timing count. Error bars show 3x replicate queries of 5x different targets.*

@@

Most methods were linear extrapolations of the single chromatogram numbers shown above as expected from a simple for loop, with notable exceptions for pyteomics (and thus the mzMLb format), the 2D peak method of pyopenms, and DuckDB (Figure 3). Pyteomics and pyopenms both had significant overhead upon initial load that resulted in faster query times afterward that performed better than a 1:1 extrapolation would expect, with pyopenms matching SQLite's speed after 10 chromatograms and Parquet's speed after 30. Interestingly, the methods that had a best-fit linear slope less than 1:1 also all had exponential fits, with performance at high chromatogram number actually worse than would be expected from a prediction fit to the timings for 1 and 3 chromatogram extractions (Supplemental figure 3).

```{r supp fig 3 multichrom}
multichrom_times <- read_csv("data/multichrom_times.csv")
multichrom_baseline_df <- multichrom_times %>%
  mutate(method=str_replace(method, "pyopenms_2DPeak", "2DPeak")) %>%
  filter(n_chroms==1) %>%
  summarise(time=mean(time), .by=method) %>%
  mutate(method=factor(method, levels=method_names))
multichrom_times %>%
  mutate(method=str_replace(method, "pyopenms_2DPeak", "2DPeak")) %>%
  mutate(method=factor(method, levels=method_names)) %>%
  ggplot(aes(x=n_chroms, y=time, color=method, fill=method)) +
  geom_abline(aes(slope=1, intercept=log10(time)), data=multichrom_baseline_df) +
  geom_smooth(method="lm", formula="y~x", color="black") +
  geom_point(position = position_dodge(width=0.1), color="black", pch=21, size=3) +
  scale_y_log10(expand=expansion(0.1)) +
  scale_x_log10(breaks=unique(multichrom_times$n_chroms), expand=expansion(0.1)) +
  scale_fill_manual(breaks = method_names, values=method_colors, 
                    aesthetics = c("fill", "color")) +
  facet_wrap(~method, scales="free_y") +
  labs(x="Number of chromatograms extracted", y="Time required (seconds)") +
  theme_bw() +
  theme(legend.position = "none")
ggsave("figures/supp_3_multichrom_facets.png", width = 6.5, height=5)
```

![](figures/supp_3_multichrom_facets.png)

*Supplemental figure 3: Data from main text Figure 3 with independent y-axes showing scatter plots of the time required to extract multiple chromatograms using various methods on logarithmic axes. Best-fit linear models have been added for each method and are shown behind triplicate timing measurements, while the black line shows a 1:1 slope intercepting the first data point to show nonlinear behavior with additional chromatograms. Transparent intervals around each best-fit line show a single standard error of the mean.*

We also explored whether database queries could be improved via the use of either a unified query (single SQL statement with multiple OR clauses for each ion's *m/z* range) or a non-equi join between a peak table with *m/z* minimum and maximum columns (Supplemental Figure 4). SQLite and Parquet performed ~3-5 times faster with the unified query than with the loop method despite the necessity of and additional processing step for the looped query to correctly assign each data point to its original peak information. The opposite was true for DuckDB likely due to its optimized reader, with the unified query consistently outperformed by the non-equi join when 100 chromatograms were extracted.

```{r supp fig 4 dbloopjoin}
type_labels = c("<span style='color:#a41118'>For loop</span>", 
               "<span style='color:#f4bb23'>Unified query</span>",
               "<span style='color:#147C8F'>Non-equi join</span>")
read_csv("data/dbloopjoinuni_times.csv") %>%
  separate(method, into = c("method", "type")) %>%
  mutate(type=case_when(
    type=="join"~"Non-equi join",
    type=="loop"~"For loop",
    type=="unified"~"Unified query"
  )) %>%
  mutate(type=factor(type, levels=c("For loop", "Unified query", "Non-equi join"),
                     labels=type_labels)) %>%
  mutate(method=factor(method, levels=c("sqlite", "duckdb", "parquet"),
                       labels=c("SQLite", "DuckDB", "Parquet"))) %>%
  ggplot(aes(x=n_chroms, y=time, fill=type)) +
  geom_point(position=position_dodge(width = 0.1), color="black", size=3, pch=21) +
  facet_wrap(~method, nrow=1) +
  scale_y_log10() +
  scale_x_log10(breaks=c(1, 3, 10, 30, 100)) +
  scale_fill_manual(breaks=type_labels, values = c("#a41118", "#f4bb23", "#147C8F")) +
  labs(x="Number of chromatograms", y="Time required (seconds)", fill=NULL) +
  theme_bw() +
  theme(legend.position = "inside", legend.justification.inside = c(0.5, 0.95),
        legend.background = element_rect(color='black'),
        legend.text=element_markdown(),
        plot.title.position = "plot", 
        plot.caption.position =  "plot")
ggsave("figures/supp_4_dbloopjoinuni.png", width = 6.5, height=5)
```

![](figures/supp_4_dbloopjoinuni.png)

*Supplemental figure 4: Scatter showing time required to extract chromatograms from SQLite, DuckDB, and Parquet databases using different methods denoted by color. Ten replicates are shown for each data point.*

@@

Certain methods are clear winners for individual use cases but all of the existing formats failed to perform well at the full suite of exploratory data analysis tasks we attempted. Scan extraction is perhaps the most widely used query, especially for proteomics, and as a result has been extensively optimized. Here, MZA was blazingly fast thanks to the decision to index in the HDF5 file by scan number. The mzML files used here did have a precompiled index that should have made scan extraction highly efficient but this appeared to be mitigated for pyopenms and pyteomics due to their long initialization times, though pymzml performed very well at this task and the other two mzML methods were much faster after initialization (Supplemental Figure 1).
